We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-02-21 paper/paper.tex
edit

In the case of M31, observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the 
total poulation of $27$ satellites \citep{2013ApJ...766..120C,
2013Natur.493...62I}, although its brightest satellites
do not show any special feature.
All published studies agree on the point that the global satellite distribution of
the 15 to 27 brightest satellites are consistent with 
spherically symmetric distribution and easy to reproduce in LCDM
simulations \citep{2006AJ....131.1405K,
  2007MNRAS.374.1125M, 2013ApJ...766..120C}.    
Table \ref{table:M31} and Table \ref{table:MW} summarize the story we
have just told, including the results from this paper using the
methodolgy we describe in the upcoming sections.

\begin{table*}
\centering
\begin{tabular}{|p{3.0cm}|p{6cm}| p{4.5cm}| c|}\hline
Reference & Target Measurement & Parent Sample & Probability ($\%$)\\\hline
\text{\cite{2006AJ....131.1405K}} & RMS height in 15 brightest
satellites & Monte Carlo satellite distributions with a power law
radial distribution & $87-99$\\
\text{\cite{2007MNRAS.374.1125M}} & Triaxiality and RMS height in 16 brightest satellites & Monte Carlo from an spherical power law radial
distribution & $17$\\
\text{\cite{2013ApJ...766..120C}}& RMS height in M31's brightest
satellites & Monte Carlo randomized satellite distribution & High\\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 24 halos from selected pairs in a cosmological N-body Dark Matter only simulation ($\sim
10^{6}$\Msun particle mass resolution)& 56 \\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 20 halos from selected pairs in a cosmological N-body hydro simulation ($\sim
10^{6}$\Msun particle mass resolution)& 27 \\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 12 high resolution DM only N-body simulations ($\sim
10^{5}$\Msun particle mass resolution) of halo pairs & 37 \\
\hline
\end{tabular}
\caption{Probability results derived in previous publications for the
  triaxiality and/or root mean squared (RMS) height of M31 satellites.
\label{table:M31}}
\end{table*}

\begin{table*}
\centering
\begin{tabular}{|p{4.0cm}|p{4.5cm}| p{5.5cm}| c|}\hline
Reference & Target Measurement & Parent Sample & Probability ($\%$)\\\hline
\text{\cite{2005A&A...431..517K}} & RMS height in 11 classical
satellites & Monte Carlo from an spherical power law
radial distribution. & $<0.5$ \\
\text{\cite{2005MNRAS.363..146L}} & Triaxiality of 11 classical
satellites & 6 high resolution DM only N-body simulations ($\sim10^5$
\Msun\ particle mass resolution). & High\\ 
\text{\cite{2005ApJ...629..219Z}} & Triaxiality and disk height in 
11 classical satellites & 3 high resolution DM only N-body
simulations. ($\sim10^6$ \Msun\ particle mass resolution) & 2 \\
\text{\cite{2007MNRAS.374.1125M}} & Triaxiality and RMS height in 11-13 brightest satellites & Monte Carlo from a halo triaxiality distribution from LCDM
simulations & $<0.5$\\
\text{\cite{2009MNRAS.399..550L}}& Triaxiality in 11 classical satellites & 436 halos from a
cosmological N-body simulation ($1.3\times 10^{8}$\Msun\ particle mass)
&  High \\
\text{\cite{2013MNRAS.429..725S}}& $c/a$ ratio in 12 brightest
satellites & 6 High resolution DM only N-body simulations ($\sim
10^3$\Msun particle mass resolution) of individual halos & Low \\
\text{\cite{2013MNRAS.429.1502W}}& $c/a$ ratio or RMS height in 11 brightest
satellites & 1686 halos from a cosmological DM only N-body simulation
($\sim 10^6$\Msun particle mass resolution) & 6 or 13 \\
\text{\cite{2014ApJ...789L..24P}}& $c/a$ and $b/a$ ratio in 11
classical satellites & 48 high resoltion DM only N-body simulations
($\sim 10^{5}$\Msun particle mass resolution) of both halo pairs and
isolated halos & 0.77\\
\text{\cite{2016MNRAS.457.1931S}}& $c/a$ ratio in 11 classical satellites & 12
high resolution Hydro simulation of halo pairs ($\sim 10^{4}$\Msun particle
mass resolution) & High\\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 24 halos from selected pairs in a cosmological N-body
Dark Matter only simulation ($\sim 10^{6}$\Msun particle mass resolution)& 3\\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 20 halos from selected pairs in acosmological N-body
hydro simulation ($\sim 10^{6}$\Msun particle mass resolution)& 0.6 \\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 12 high resolution DM only N-body simulations ($\sim
10^{5}$\Msun particle mass resolution) of halo pairs & 0.2 \\
\hline
\end{tabular}
\caption{Same as Table \ref{table:M31} for the MW satellites.
\label{table:MW}}
\end{table*}

@@

In this paper, we surmount the first difficulty by setting as a point of reference a
spherical satellite distribution and not LCDM simulations.

This spherical distribution is built from the data itself
(observational or simulated) by randomizing its angular position
around the central galaxy and keeping its radial distance fixed
\citep{2017AN....338..854P}. 
We characterize each satellite distribution with its inertia
tensor and summarize the results with the scalars describing the width
of the distribution and its axis ratios.

We handle the second difficuly by building an explicit analytic
probability distribution that is compatible with the simulated data.
We then use the simulations to estimate the parameters in this
distribution to later use it as a parent sample to generate any
desired number of samples that are by construction statistically
compatible with simulations.

To summarize, simulations help us first to construct an explicit probability
distribution for deviations from sphericity for a given LCDM simulated
sample and then, using this distribution, we compute the odds of finding a satellite
distribution that deviates from sphericity as much as the LG does.

>>>>

In the case of M31, observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the 
total poulation of $27$ satellites \citep{2013ApJ...766..120C,
2013Natur.493...62I}, although its brightest satellites
do not show any special feature.
All published studies agree on the point that the global satellite distribution of
the 15 to 27 brightest satellites are consistent with 
spherically symmetric distribution and easy to reproduce in LCDM
simulations \citep{2006AJ....131.1405K,
  2007MNRAS.374.1125M, 2013ApJ...766..120C}.    
Table \ref{table:M31} and Table \ref{table:MW} summarize some more
details on the results we have just mentioned.
Those tables also include the results from this paper using the
methodolgy we describe in the upcoming sections.

\begin{table*}
\centering
\begin{tabular}{|p{4.0cm}|p{4.5cm}| p{5.5cm}| c|}\hline
Reference & Target Measurement & Parent Simulation & Probability ($\%$)\\\hline
\text{\cite{2006AJ....131.1405K}} & RMS height in 15 brightest
satellites & Monte Carlo satellite distributions with a power law
radial distribution & $87-99$\\
\text{\cite{2007MNRAS.374.1125M}} & Triaxiality and RMS height in 16 brightest satellites & Monte Carlo from an spherical power law radial
distribution & $17$\\
\text{\cite{2013ApJ...766..120C}}& RMS height in the 27 brightest
satellites & Monte Carlo randomized satellite distribution & High\\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 24 halos from selected pairs in a cosmological N-body Dark Matter only simulation ($\sim
10^{6}$\Msun particle mass resolution)& 56 \\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 20 halos from selected pairs in a cosmological N-body hydro simulation ($\sim
10^{6}$\Msun particle mass resolution)& 27 \\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 12 high resolution DM only N-body simulations ($\sim
10^{5}$\Msun particle mass resolution) of halo pairs & 37 \\
\hline
\end{tabular}
\caption{Probability results derived in previous publications for the
  triaxiality and/or root mean squared (RMS) height of M31 satellites.
\label{table:M31}}
\end{table*}

\begin{table*}
\centering
\begin{tabular}{|p{4.0cm}|p{4.5cm}| p{5.5cm}| c|}\hline
Reference & Target Measurement & Parent Simulation & Probability ($\%$)\\\hline
\text{\cite{2005A&A...431..517K}} & RMS height in 11 classical
satellites & Monte Carlo from an spherical power law
radial distribution. & $<0.5$ \\
\text{\cite{2005MNRAS.363..146L}} & Triaxiality of 11 classical
satellites & 6 high resolution DM only N-body simulations ($\sim10^5$
\Msun\ particle mass resolution). & High\\ 
\text{\cite{2005ApJ...629..219Z}} & Triaxiality and disk height in 
11 classical satellites & 3 high resolution DM only N-body
simulations. ($\sim10^6$ \Msun\ particle mass resolution) & 2 \\
\text{\cite{2007MNRAS.374.1125M}} & Triaxiality and RMS height in 11-13 brightest satellites & Monte Carlo from a halo triaxiality distribution from LCDM
simulations & $<0.5$\\
\text{\cite{2009MNRAS.399..550L}}& Triaxiality in 11 classical satellites & 436 halos from a
cosmological N-body simulation ($1.3\times 10^{8}$\Msun\ particle mass)
&  High \\
\text{\cite{2013MNRAS.429..725S}}& $c/a$ ratio in 12 brightest
satellites & 6 High resolution DM only N-body simulations ($\sim
10^3$\Msun particle mass resolution) of individual halos & Low \\
\text{\cite{2013MNRAS.429.1502W}}& $c/a$ ratio or RMS height in 11 brightest
satellites & 1686 halos from a cosmological DM only N-body simulation
($\sim 10^6$\Msun particle mass resolution) & 6 or 13 \\
\text{\cite{2014ApJ...789L..24P}}& $c/a$ and $b/a$ ratio in 11
classical satellites & 48 high resoltion DM only N-body simulations
($\sim 10^{5}$\Msun particle mass resolution) of both halo pairs and
isolated halos & 0.77\\
\text{\cite{2016MNRAS.457.1931S}}& $c/a$ ratio in 11 classical satellites & 12
high resolution Hydro simulation of halo pairs ($\sim 10^{4}$\Msun particle
mass resolution) & High\\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 24 halos from selected pairs in a cosmological N-body
Dark Matter only simulation ($\sim 10^{6}$\Msun particle mass resolution)& 3\\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 20 halos from selected pairs in acosmological N-body
hydro simulation ($\sim 10^{6}$\Msun particle mass resolution)& 0.6 \\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 12 high resolution DM only N-body simulations ($\sim
10^{5}$\Msun particle mass resolution) of halo pairs & 0.2 \\
\hline
\end{tabular}
\caption{Same as Table \ref{table:M31} for the MW satellites.
\label{table:MW}}
\end{table*}

@@

In this paper we surmount the first difficulty by setting as direct 
point of reference an spherical satellite distribution and not the
simulations themselves.
The spherical distribution is built from the data itself
(observational or simulated) by randomizing its angular position
around the central galaxy and keeping its radial distance fixed
\citep{2017AN....338..854P}. 
We then characterize each satellite distribution in terms of the
scalars describing its deviation from the spherical distribution.
We handle the second difficulty by building an explicit analytic
probability distribution of the deviations from asphericity; this
solves the problem of having a common reference point to compare
simulations. 
We then use different simulations to estimate the parameters in this
distribution to later use it as a parent sample to generate any
desired number of samples that are by construction statistically
compatible with simulations, overcoming the problem of having a small
number of systems in the parent simulations.

To summarize, we start by using asphericity to characterize in the
same way simulations and observations. 
Then, we build an explicit probability distribution for the asphericity and
use simulations to estimate the parameters in this
distribution. 
Finally, we use these distributions to generate large number of
samples and directly estimate the number of systems meeting a desired
set of criteria.
